Analysis of Motor Vehicle Collision Causes in Toronto from 2006 to 2023

### 2024-09-27 paper/paper.qmd
major update on paper

---
title: "Analysis of Motor Vehicle Collision Causes in Toronto from 2006 to 2023"
subtitle: "Exploring the Interplay of Human Behavior and Environmental Factors in Predicting Accident Severity and High-Risk Scenarios"
author: 
  - Jiwon Choi
thanks: "Code and data are available at: https://github.com/jwonc4602/Analysis-of-Motor-Vehicle-Collision-Causes."
date: today
date-format: long
abstract: "This paper analyzes motor vehicle collisions in Toronto from 2006 to 2023, focusing on the interplay between human behavior and environmental factors. The study examines how behaviors such as speeding, aggressive driving, alcohol consumption, and disability contribute to accidents alongside conditions like visibility, lighting, and road surface. Findings reveal that while environmental factors can exacerbate risks, human actions, particularly aggressive driving and alcohol use, are the most significant contributors to severe collisions. These insights are critical for developing predictive models and preventive strategies that can enhance urban road safety by addressing both driver behavior and infrastructure improvements."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Motor vehicle collisions remain one of the leading causes of injury and fatality in urban environments, prompting ongoing research into how these incidents can be mitigated. In a rapidly growing city like Toronto, understanding the causes of traffic accidents is essential for improving road safety and informing policy. This paper investigates traffic collisions in Toronto over a 17-year span, analyzing the combined effects of human behavior and environmental conditions to determine which factors most significantly contribute to the frequency and severity of accidents.

Despite extensive research on individual factors like speeding or poor road conditions, there remains a gap in understanding how human behaviors interact with environmental conditions to create high-risk scenarios for accidents. Existing studies tend to focus on either human or environmental factors in isolation, missing the crucial interplay between these elements. This research aims to fill that gap by examining both human and environmental factors in tandem, providing a more holistic view of traffic accident causes in an urban context.

Using data from Toronto Police Services, this study explores the role of speeding, aggressive driving, alcohol consumption, and disability alongside environmental factors such as visibility, lighting, traffic control, and road surface conditions. The findings reveal that aggressive driving and alcohol use, especially under favorable road conditions, are leading contributors to collisions. The study also identifies high-risk combinations, such as aggressive driving on dry roads under clear visibility, which result in the highest number of accidents. These results are important because they offer a nuanced understanding of traffic safety and suggest that targeted interventions, such as stricter penalties for reckless driving or improved traffic control in key areas, could significantly reduce accidents.

The remainder of this paper is structured as follows. Section 2 provides a detailed description of the data and methodology. Section 3 presents the analysis of human and environmental factors contributing to traffic collisions. Section 4 discusses the results, highlighting high-risk scenarios and potential interventions. Finally, Section 5 addresses the limitations of the current study and outlines directions for future research, particularly the need to account for the growing presence of autonomous driving technologies in collision analysis.

@@

This traffic accident data consists of 18,957 entries from a citywide accident survey, capturing detailed information about various factors involved in traffic incidents. The raw dataset includes 50 variables, ranging from geographic data, such as the street where the accident occurred, to behavioral data, such as driver condition and alcohol involvement. This dataset records accidents over multiple years and provides insights into environmental conditions, road types, and the actions of drivers and pedestrians at the time of the incidents. The goal of this dataset is to provide a comprehensive view of traffic accidents for research and policy analysis related to urban traffic safety and accident prevention.

The cleaned data is divided into two datasets, and the first one focuses on human factors related to traffic collisions, specifically the behaviors and conditions of drivers that may have contributed to accidents. The four main categories—'speeding', 'ag_driv', 'alcohol', and 'disability'—were selected because these are well-known, critical factors that significantly increase the risk of traffic accidents. Each factor is evaluated as either "None" or "Yes", indicating the absence or presence of that factor in the collision. These categories were chosen because they represent preventable human actions or conditions that are highly correlated with the likelihood and severity of traffic incidents. (see @tbl-cleaned-human-factors-data)

1. **speeding**: Refers to collisions where speeding was a contributing factor. Speeding reduces a driver's ability to react quickly to road conditions or obstacles, and it increases the severity of crashes when they occur, making it one of the primary causes of traffic collisions.

2. **ag_driv**: This category examines whether aggressive behavior, such as reckless driving, or distracted driving (like texting or talking on the phone) was involved in the collision. Aggressive and distracted driving are known to impair judgment and reaction time, leading to accidents that could otherwise have been avoided.

3. **alcohol**: Evaluates if alcohol was a factor in the collision. Alcohol impairs motor skills and decision-making, and is a well-documented contributor to severe and fatal traffic accidents. This category is crucial for understanding the role of substance use in driver behavior.

4. **disability**: Assesses if any medical or physical disability played a role in the collision. Disabilities, whether temporary or permanent, can affect a driver's ability to control the vehicle or respond to changing road conditions, making this an important factor to consider in the context of traffic safety.

These categories were selected because they represent key human factors that have a direct impact on traffic safety, and addressing them can help reduce both the frequency and severity of accidents.

@@

# Line plot to show trends over time
ggplot(trend_data, aes(x = year, y = count, color = factor)) +
  geom_line(linewidth = 0.8) +
  theme_minimal() +
  labs(title = "Trends in Human Factors Over Time",
       x = "Year",
       y = "Count of Collisions",
       color = "Human Factor")
```

The second dataset focuses on the environmental factors influencing traffic collisions, capturing the conditions that shape driver behavior and contribute to accidents. These factors, including traffic control, visibility, lighting, and road surface conditions, play a vital role in either increasing or reducing the risk of collisions. The following modifications were made to simplify and group these categories, making the data more consistent and easier to analyze. The merged categories provide clearer insights into how these factors affect accidents and highlight areas for safety improvements. (see @tbl-cleaned-environmental-factors-data)

1. **traffic_control_type**: This category was simplified to include "No Control," "Traffic Signal," and "Others." The "Others" category consolidates less frequent types of traffic control, such as "Stop Sign," "Pedestrian Crossover," and "School Guard." This grouping reflects the most significant distinctions in how traffic is regulated, emphasizing scenarios where the presence or absence of formal traffic control could impact driver behavior. By focusing on the broad types of control, this category helps assess how different levels of traffic management influence accident risks.

2. **visibility**: The visibility conditions were grouped into "Clear" and "Disturbed." The "Disturbed" category includes various adverse weather conditions such as "Fog, Mist, Smoke, Dust," "Rain," "Snow," and "Strong Wind." This grouping simplifies the analysis by highlighting the key distinction between ideal and hazardous visibility conditions, which directly affect a driver’s ability to see and react to road conditions. Poor visibility is a known contributor to accidents, making this distinction critical for identifying environmental hazards.

3. **light**: The lighting conditions were merged into three categories: "Dark" (which includes "Dark" and "Dark, artificial"), "Daylight" (including "Daylight" and "Daylight, artificial"), and "Dim Light" (covering "Dawn," "Dusk," and similar intermediate lighting conditions). These groups reflect the significant lighting scenarios that impact driver visibility. Dark and dim lighting can reduce reaction times and perception, making these categories essential for understanding how lighting conditions contribute to accidents.

4. **road_surface_condition**: Road surface conditions were consolidated into "Dry," "Wet," and "Others." The "Others" category includes conditions like "Ice," "Loose Snow," and "Slush," which can severely reduce traction. Grouping less common conditions into "Others" helps streamline the analysis by focusing on the primary road conditions that affect vehicle control. Wet or slippery surfaces are known to increase accident risks, making this distinction important for understanding how road conditions influence collisions.

These categories were selected and modified to prioritize the most critical environmental factors that affect traffic safety. By simplifying and grouping related variables, the data becomes more interpretable, allowing for a clearer focus on how environmental factors like traffic control, visibility, lighting, and road conditions contribute to motor vehicle collisions. This approach not only aids in identifying preventable causes but also supports efforts to improve road safety and reduce collision rates through targeted interventions.

@@

Lastly, the **acclass** category classifies the type and severity of traffic collisions with labels such as "Fatal," "Non-Fatal Injury," and "None or Others." The original classifications of "None" and "Property Damage Only" were merged into the "None or Others" category to streamline the data, as both involve accidents without human injury. This consolidation was taken to simplify the analysis by focusing on the human impact of accidents. "Fatal" represents collisions resulting in severe injury, and "Non-Fatal Injury" indicates incidents causing injury but no fatalities. By merging and refining these categories, the focus shifts toward understanding the human consequences of collisions, enabling a more targeted approach to reducing severe injuries and fatalities.

@@

Upon further review of the heatmap, the top three combinations of environmental and human factors are evident, and I will revise the analysis accordingly.

### High-Risk Scenarios for Road Accidents

The heatmap highlights the intricate relationship between environmental conditions—such as traffic control, visibility, lighting, and road surface—and human factors like speeding, aggressive driving, alcohol use, and disability. This visualization reveals clear trends where specific combinations of these factors lead to the highest rates of road accidents.

The highest-risk combination occurs when **traffic signals are clear, daylight is present, the road is dry, and drivers exhibit aggressive driving behavior**. This combination results in the largest number of accidents. Even though environmental conditions are favorable with clear daylight and dry roads, aggressive driving significantly increases the accident rate. This indicates that aggressive behaviors, such as tailgating or reckless overtaking, are extremely dangerous even in otherwise safe driving conditions. To mitigate these risks, more targeted enforcement of speed limits and penalties for aggressive driving may be necessary in areas where accidents are high, despite good road conditions.

A second high-risk combination involves **no traffic control, clear conditions, daylight, and dry roads, combined with aggressive driving and alcohol use**. This combination results in high accident counts, emphasizing how dangerous aggressive driving becomes when compounded by alcohol consumption, even in seemingly safe environmental conditions. The absence of traffic control in these situations suggests that drivers may feel a false sense of security, which encourages reckless behavior. Enhancing traffic control measures, such as adding stop signs, traffic lights, or roundabouts in high-accident zones, could reduce incidents in these cases.

The third-highest combination includes **no traffic control, disturbed visibility, daylight, and wet road conditions** with **no involvement of impaired behaviors**. This scenario shows how environmental factors like wet roads and poor visibility increase the risk of accidents, even when human behavior is not a contributing factor. The lack of traffic control adds further risk. Addressing this issue through better traffic control measures and improving road surface conditions, such as better drainage systems or warning signs in areas prone to heavy rainfall, can help reduce accidents.

The data reveals that while environmental factors such as wet roads, disturbed visibility, and lack of traffic control are significant contributors to accidents, aggressive driving and alcohol use are key human behaviors that exacerbate risks, even in favorable environmental conditions. Focusing on infrastructure improvements, such as enhancing traffic control in key areas and stricter enforcement against aggressive driving and alcohol-related offenses, could significantly reduce accident rates in the most hazardous scenarios.

@@

## Can Environmental and Human Factors Serve as Early Warning Indicators?
With the advancement of data-driven technologies, it is possible to predict the severity of motor vehicle accidents based on a combination of human behaviors and environmental conditions. This discussion focuses on how predictive models can be developed to identify the likelihood of severe accidents using factors like speeding, alcohol consumption, poor road conditions, and low visibility.

For example, we could create a system that collects real-time environmental data (e.g., weather, road surface conditions) and cross-references it with information on traffic behaviors (e.g., average speeding rates) to predict whether an area is at high risk for serious accidents. This kind of early warning system could be integrated into navigation apps or smart cars to alert drivers of heightened risks and potentially prevent accidents before they occur.

Predictive analytics can highlight the most dangerous combinations of factors, such as speeding in rainy weather or aggressive driving under poor visibility. In addition, machine learning algorithms can continuously improve these predictions by incorporating accident history data, making the model more robust over time. This offers the potential to shift from reactive to proactive accident prevention, reducing the burden on healthcare systems and increasing road safety.

@@

One notable limitation of this analysis is that the dataset only contains information up until 2023. While this provides a substantial historical perspective on traffic accidents in Toronto, it may not fully capture recent changes in road usage patterns, traffic enforcement, or driver behaviors that have emerged post-2023. For example, the rise of new driving technologies, ride-sharing services, and changes in traffic regulations could have a significant impact on accident rates, but these factors are not reflected in the current dataset. As such, the findings based on this data may not completely align with more recent trends, limiting the relevance of the analysis for current or future policy decisions.

While the dataset used in this analysis ends in 2023, a key area for future exploration involves the broader adoption of autonomous or auto-drive technologies, which are not yet accounted for in the data but could significantly impact traffic accident dynamics. As these technologies become more prevalent, understanding their effect on traffic safety, accident rates, and accident severity will be critical. The absence of this data limits our ability to predict how autonomous driving might influence future accident patterns. However, from the figure "fig-trend-over-time," we observe that collisions involving aggressive driving (ag_driv) have decreased over time, which may be indicative of increased use of autonomous or driver-assist technologies that mitigate reckless behavior. Future research should incorporate data on autonomous vehicles and their interaction with human-driven cars to better assess how these advancements influence safety and accident risks. This could lead to the development of predictive models that combine human and automated driving behaviors, aiding policymakers and urban planners in shaping infrastructure and safety regulations for a future where both types of vehicles coexist.

>>>>

---
title: "Analysis of Motor Vehicle Collision Causes in Toronto from 2006 to 2023"
subtitle: "Exploring the Interplay of Human Behavior and Environmental Factors in Predicting Accident Severity and High-Risk Scenarios"
author: 
  - Jiwon Choi
thanks: "Code and data are available at: https://github.com/jwonc4602/Analysis-of-Motor-Vehicle-Collision-Causes."
date: today
date-format: long
abstract: "This study analyzes motor vehicle collisions in Toronto from 2006 to 2023, focusing on the interaction between human behaviors—such as speeding, aggressive driving, and alcohol use—and environmental conditions like visibility and road surfaces. The findings show that while environmental factors increase risks, aggressive driving and alcohol use are the strongest predictors of severe accidents. These results highlight the need for targeted interventions in both driver behavior and infrastructure to improve urban traffic safety."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Motor vehicle collisions are a leading cause of injury and fatality in urban areas, driving ongoing efforts to improve road safety. In a rapidly expanding city like Toronto, understanding the causes of traffic accidents is critical for shaping effective policies. This paper examines traffic collisions in Toronto over a 17-year period, focusing on how human behavior and environmental conditions together influence the frequency and severity of accidents.

While much research has been done on factors such as speeding or poor road conditions, there is still a lack of understanding regarding the interaction between human behaviors and environmental conditions in creating high-risk situations. Most studies focus on these factors separately, missing how they combine to increase accident risks. This research addresses that gap by analyzing both human and environmental factors simultaneously, offering a more complete view of what drives traffic accidents in an urban setting.

Using data from Toronto Police Services, this study analyzes the role of human factors like speeding, aggressive driving, alcohol use, and disability alongside environmental factors such as visibility, lighting, traffic control, and road conditions. The results show that aggressive driving and alcohol use, even under good road conditions, are major contributors to accidents. High-risk scenarios, like aggressive driving on dry roads with clear visibility, emerge as leading causes of collisions. These findings suggest targeted actions, such as stricter enforcement against reckless driving and improved traffic controls in key areas, could help reduce accident rates.

The rest of the paper is organized as follows. Section 2 outlines the data and methodology. Section 3 presents the analysis of factors contributing to traffic collisions. Section 4 discusses the results and potential interventions. Section 5 addresses the study’s limitations and proposes future research directions, particularly the need to consider the growing role of autonomous driving technologies in accident analysis.

@@

This traffic accident dataset contains 18,957 entries from a citywide survey, detailing various factors involved in incidents. It includes 50 variables covering geographic data, such as accident locations, and behavioral data, like driver conditions and alcohol involvement. The dataset spans several years and captures environmental conditions, road types, and the actions of drivers and pedestrians during accidents. It aims to provide a comprehensive view for research and policy efforts related to urban traffic safety and accident prevention.

The cleaned data is split into two sets. The first focuses on human factors, particularly driver behaviors and conditions that may have contributed to accidents. Four key categories—'speeding,' 'ag_driv,' 'alcohol,' and 'disability'—were selected due to their known impact on traffic accidents. Each is recorded as either "None" or "Yes," indicating whether that factor was present in the incident. These categories focus on preventable human actions or conditions that are strongly linked to traffic risks.

1. **Speeding**: Tracks whether speeding was a factor in the collision. Speeding reduces reaction time and increases crash severity, making it a leading cause of accidents.

2. **Aggressive Driving (ag_driv)**: Refers to reckless or distracted behaviors, such as texting while driving. These actions impair judgment and reaction time, significantly increasing accident risk.

3. **Alcohol**: Captures whether alcohol was involved in the collision. Alcohol impairs motor skills and decision-making, contributing to many severe and fatal accidents.

4. **Disability**: Notes if a medical or physical disability affected the driver's ability to control the vehicle. Disabilities can impair a driver's capacity to respond to road conditions, making this a key safety factor.

These factors are critical to understanding the human behaviors that increase traffic accident risks and can inform strategies to reduce both the frequency and severity of collisions.

@@

# Line plot to show trends over time
ggplot(trend_data, aes(x = year, y = count, color = factor)) +
  geom_line(linewidth = 0.8) +
  theme_minimal() +
  labs(title = "Trends in Human Factors Over Time",
       x = "Year",
       y = "Count of Collisions",
       color = "Human Factor")
```
The second dataset focuses on environmental factors that influence traffic collisions, including conditions such as traffic control, visibility, lighting, and road surface. These elements shape driver behavior and impact accident risk. To simplify analysis and improve consistency, categories were consolidated, providing clearer insights into how these factors affect accidents and where safety improvements are needed. (see @tbl-cleaned-environmental-factors-data)

1. **Traffic Control Type**: This was simplified into "No Control," "Traffic Signal," and "Others." The "Others" group combines less common traffic controls like stop signs and pedestrian crossovers. This distinction highlights the impact of varying levels of traffic regulation on driver behavior, focusing on scenarios where traffic control presence may influence accident risks.

2. **Visibility**: Conditions were grouped into "Clear" and "Disturbed," with "Disturbed" covering adverse weather such as fog, rain, and snow. This division highlights the contrast between ideal and hazardous visibility, which directly affects a driver's ability to navigate safely. Poor visibility is a known factor in accidents, making this an essential category for identifying environmental hazards.

3. **Light**: Lighting conditions were combined into "Dark," "Daylight," and "Dim Light." These reflect the most relevant lighting environments, with "Dark" including both natural and artificial lighting, and "Dim Light" covering intermediate conditions like dawn and dusk. Lighting plays a crucial role in visibility and reaction times, and this grouping helps assess its role in accidents.

4. **Road Surface Condition**: This was grouped into "Dry," "Wet," and "Others." The "Others" category includes ice, snow, and slush, conditions that greatly reduce vehicle traction. Focusing on major surface conditions helps clarify how road conditions influence accident risk, especially in wet or slippery environments.

These modifications prioritize the key environmental factors affecting traffic safety. By simplifying and grouping variables, the dataset becomes easier to interpret, allowing a clearer understanding of how conditions like traffic control, visibility, lighting, and road surfaces contribute to collisions. This approach supports targeted interventions aimed at improving road safety and reducing accident rates.

@@

The **acclass** category classifies the severity of traffic collisions into "Fatal," "Non-Fatal Injury," and "None or Others." The original categories "None" and "Property Damage Only" were combined into "None or Others" since both involve accidents without human injury. This consolidation simplifies the analysis, focusing on accidents with a direct human impact. "Fatal" refers to collisions causing severe injury or death, while "Non-Fatal Injury" includes incidents with injuries but no fatalities. By merging and refining these categories, the data emphasizes the human consequences of accidents, helping target efforts to reduce severe injuries and fatalities.

@@

The heatmap shows the relationship between environmental factors—traffic control, visibility, lighting, and road surface—and human behaviors like speeding, aggressive driving, alcohol use, and disability. It highlights clear patterns where certain combinations lead to higher accident rates.

The most dangerous combination occurs when **traffic signals are present, daylight is clear, the road is dry, and aggressive driving is involved**. Even in ideal environmental conditions, aggressive driving leads to the highest number of accidents. This suggests that behaviors like tailgating and reckless overtaking pose significant risks, even on dry roads with good visibility. To reduce accidents in such areas, stricter enforcement of speed limits and penalties for aggressive driving may be necessary.

Another high-risk combination is **no traffic control, clear conditions, daylight, dry roads, and aggressive driving coupled with alcohol use**. The lack of traffic control in this scenario appears to give drivers a false sense of safety, leading to reckless behavior. Installing traffic controls, like stop signs or roundabouts, in areas with high accident rates could help reduce these incidents.

A third combination showing increased risk involves **no traffic control, poor visibility, daylight, and wet roads**, but **no involvement of risky behaviors**. This highlights how wet roads and low visibility increase accident risks, even without human error. Improving road infrastructure, such as adding drainage systems or warning signs in areas prone to heavy rainfall, and introducing traffic controls could help prevent accidents under these conditions.

The data shows that while environmental factors like wet roads and poor visibility are important, aggressive driving and alcohol use are key behaviors that amplify risk, even in favorable conditions. Focusing on improving infrastructure and stricter enforcement against aggressive driving and alcohol-related offenses could significantly lower accident rates in these high-risk scenarios.

@@

## Can Environmental and Human Factors Serve as Early Warning Indicators?

With advances in data-driven technologies, it's now possible to predict the severity of motor vehicle accidents by analyzing a combination of human behaviors and environmental conditions. This section focuses on how predictive models can be developed to assess the likelihood of severe accidents using factors such as speeding, alcohol use, poor road conditions, and low visibility.

For instance, a system could gather real-time environmental data, such as weather and road surface conditions, and combine it with information on traffic behaviors, like speeding patterns, to determine if an area is at high risk for serious accidents. This type of early warning system could be integrated into navigation apps or smart cars to alert drivers to potential dangers and help prevent accidents before they occur.

Predictive analytics can identify the most dangerous combinations of factors, such as speeding during rainy weather or aggressive driving in poor visibility. Machine learning algorithms could continuously improve these models by incorporating historical accident data, making the predictions more accurate over time. This approach shifts accident prevention from a reactive to a proactive strategy, which could reduce the strain on healthcare systems and improve overall road safety.

@@

One limitation of this analysis is that the dataset only includes data up to 2023. While this offers a strong historical perspective on traffic accidents in Toronto, it does not account for more recent shifts in road usage, traffic enforcement, or driver behavior that may have emerged after 2023. For example, new driving technologies, ride-sharing services, and updated traffic regulations could significantly influence accident rates, but these factors are not reflected in the current data. As a result, the findings may not fully align with recent trends, limiting their relevance for current and future policy decisions.

A key area for future research involves the growing use of autonomous or driver-assist technologies, which are not yet covered by the dataset but could have a major impact on traffic accident patterns. As these technologies become more common, understanding their effect on traffic safety and accident severity will be essential. The lack of data on autonomous vehicles limits our ability to predict how these innovations might change accident trends. However, from the figure @fig-trend-over-time, we see that accidents involving aggressive driving have decreased, possibly due to the rise of driver-assist features that reduce reckless behavior. Future studies should incorporate data on autonomous vehicles and their interactions with human-driven cars to better understand how these advancements affect safety. This could help develop predictive models that account for both human and automated driving behaviors, providing valuable insights for policymakers and urban planners as they adapt infrastructure and safety regulations for a future where both types of vehicles share the road.
